Tensor denoising and completion based on ordinal observations

### 2020-02-18 ICML/submit/supplements/latexsource/ordinalTv3.tex
Add files via upload

\begin{icmlauthorlist}
\icmlauthor{Chanwoo Lee}{equal,to}
\icmlauthor{Miaoyan Wang}{equal,to}
\end{icmlauthorlist}

\icmlaffiliation{to}{Department of Statistics, University of Wisconsin at Madison, USA}

\icmlcorrespondingauthor{Miaoyan Wang}{miaoyan.wang@wisc.edu.com}

% You may provide any keywords that you
% find helpful for describing your paper; these are used to populate
% the "keywords" metadata in the PDF but will not be shown in the document
\icmlkeywords{Higher-order tensors, ordinal observation, tensor decomposition, tensor completion}

@@

\section{Introduction}
Multidimensional arrays, a.k.a.\ tensors, arise in a variety of applications including recommendation systems~\citep{baltrunas2011incarmusic}, social networks~\citep{nickel2011three}, genomics~\citep{hore2016tensor}, and neuroimaging~\citep{zhou2013tensor}. There is a growing need to develop general methods for analyzing these noisy, high-dimensional datasets that can handle two main problems. The problem of tensor denoising -- which aims to recover a signal tensor from its noisy entries -- has gained increased attention in theory and applications~\citep{xia2019sup,zeng2019multiway}. A related problem, tensor completion, examines the minimum number of entries needed for a consistent recovery~\citep{ghadermarzy2018learning,ghadermarzy2019near}. Low-rankness is often imposed to the signal tensor, which efficiently reduces the intrinsic dimension in both problems.

@@

\section{Preliminaries}
Let $\tY\in\mathbb{R}^{d_1\times \cdots \times d_K}$ denote an order-$K$ $(d_1,\ldots,d_K)$-dimensional tensor. We use $y_\omega$ to denote the tensor entry indexed by $\omega$, where $\omega\in[d_1]\times\cdots\times[d_K]$.  The Frobenius norm of $\tY$ is defined as $\FnormSize{}{\tY}=\sum_\omega y^2_\omega$ and the infinity norm of $\tY$ is defined as $\mnormSize{}{\tY}=\max_{\omega}|y_\omega|$. We use $\tY_{(k)}$ to denote the unfolded matrix of size $d_k$-by-$\prod_{i\neq k}d_k$, obtained by reshaping the tensor along the mode-$k$. The Tucker rank of $\tY$ is defined as a length-$K$ vector $\mr = (r_1,\ldots,r_K)$, where $r_k$ is the rank of matrix $\tY_{(k)}$ for all $k \in[K]$. We say that an event $A$ occurs ``with very high probability'' if $\mathbb{P}(A)$ tends to 1 faster than any polynomial of tensor dimension $d_{\min}=\min\{d_1,\ldots,d_K\} \to\infty$.

@@

\vspace{-.2cm}
\section{Model formulation and motivation}
\subsection{Observation model}
Let $\tY$ denote an order-$K$ $(d_1,\ldots,d_K)$-dimensional data tensor. Suppose the entries of $\tY$ are ordinal-valued, and the observation space consists of $L$ ordered levels, denoted by $[L]:=\{1,\ldots,L\}$. We propose a cumulative link model for the ordinal tensor $\tY=\entry{y_\omega}\in[L]^{d_1\times \cdots\times d_K}$. Specifically, assume the entries $y_\omega$ are (conditionally) independently distributed with cumulative probabilities:
\begin{equation}\label{eq:model}
\mathbb{P}(y_\omega\leq \ell)=f(b_\ell-\theta_\omega),\ \text{for all}\ \ell\in[L-1],
\end{equation}
where $\mb=(b_1,\ldots,b_{L-1})$ is a set of unknown scalars satisfying $b_1<\cdots <b_{L-1}$, $\Theta=\entry{\theta_\omega}\in\mathbb{R}^{d_1\times \cdots \times d_K}$ is a continuous-valued parameter tensor satisfying certain low-dimensional structure (to be specified later), and $f(\cdot):\mathbb{R}\mapsto[0,1]$ is a known, strictly increasing function. We refer to $\mb$ as the cut-off points and $f$ the link function.

The formulation~\eqref{eq:model} imposes an additive model to the transformed probability of cumulative categories. This modeling choice is to respect the ordering structure among the categories. For example, if we choose the inverse link $f^{-1}(x)=\log {x\over 1-x}$ to be the log odds, then the model~\eqref{eq:model} implies linear spacing between the proportional odds:
\begin{equation}\label{eq:logodd}
\log {\mathbb{P}(y_{\omega}\leq \ell) \over \mathbb{P}(y_\omega >  \ell) } - \log {\mathbb{P}(y_{\omega}\leq {\ell-1}) \over \mathbb{P}(y_\omega >  {\ell-1}) } = b_\ell-b_{\ell-1},
\end{equation}
for all tensor entries $y_\omega$. When there are only two categories in the observation space (e.g. binary tensors), the cumulative model~\eqref{eq:model} is equivalent to the usual multinomial link model. In general, however, when the number of categories $L\geq 3$, the proportional odds assumption~\eqref{eq:logodd} is more parsimonious, in that, the ordered categories can be envisaged as contiguous intervals on the continuous scale, where the points of division are exactly $b_1<\cdots <b_{L-1}$. This interpretation will be made more explicit in the next section.

\subsection{Latent-variable interpretation}\label{sec:latent}
The ordinal tensor model~\eqref{eq:model} with certain types of link $f$ has the equivalent representation as an $L$-level quantization model on $\tY=\entry{y_\omega}$:
\begin{align}\label{eq:quantization}
y_\omega&=
\begin{cases}
1,& \text{if $y^*_\omega\in(-\infty, b_1]$},\\
2,& \text{if $y^*_\omega\in(b_1, b_2]$},\\
\vdots  &\vdots\\
L,& \text{if $y^*_\omega\in(b_{L-1}, \infty)$},\\
\end{cases}
\end{align}
for all $\omega\in[d_1]\times \cdots \times [d_k]$. Here, $\tY^*=\entry{y^*_\omega}$ is a latent continuous-valued tensor following an additive noise model:
\begin{align}\label{eq:latent}
\KeepStyleUnderBrace{\tY^*}_{\text{latent continuous-valued tensor}}=\KeepStyleUnderBrace{\Theta}_{\text{signal tensor}}+\KeepStyleUnderBrace{\tE}_{\text{i.i.d.\ noise}},
\end{align}
where $\tE=\entry{\varepsilon_\omega}\in\mathbb{R}^{d_1\times \cdots \times d_K}$ is a noise tensor with i.i.d.\ entries according to distribution $\mathbb{P}(\varepsilon)$. From the viewpoint of~\eqref{eq:latent}, the parameter tensor $\Theta$ can be interpreted as the latent signal tensor prior to contamination and quantization.

@@

\begin{assumption}\label{ass:link}
The link function is assumed to satisfy:
\vspace{-.4cm}
\begin{enumerate}
\item $f(\theta)$ is strictly increasing and twice-differentiable in $\theta\in\mathbb{R}/\{0\}$.
\vspace{-.2cm}
\item $f'(\theta)$ is strictly log-concave and symmetric with respect to $\theta=0$.
\end{enumerate}
\end{assumption}

\subsection{Problem 1: Tensor denoising}~\label{sec:denoising}
The first question we aim to address is tensor denoising:

@@

The second constraint is that the entries of $\Theta$ are uniformly bounded in magnitude by a constant $\alpha \in \mathbb{R}_{+}$. In view of~\eqref{eq:latent}, we refer to $\alpha$ as the signal level. The entry-wise bound assumption is a technical condition that avoids the degeneracy in probability estimation with ordinal observations.

\subsection{Problem 2: Tensor completion}
Motivated by applications in collaborative filtering, we also consider a more general setup when only a subset of tensor entries $y_\omega$ are observed. Let $\Omega\subset[d_1]\times \cdots\times[d_K]$ denote the set of observed indices. The second question we aim to address is stated as follows:

@@

We consider the same parameter space~\eqref{eq:space} for the completion problem. In addition to the reasons mentioned in Section~\ref{sec:denoising}, the entrywise bound assumption also serves as the incoherence requirement for completion. In classical matrix completion, the incoherence is often imposed on the singular vectors. This assumption is recently relaxed for ``flat'' matrices with bounded magnitude~\cite{negahban2011estimation,cai2013max,bhaskar20151}. We adopt the same assumption for higher-order tensors.

\section{Rank-constrained M-estimator}\label{sec:theory}
We present a general treatment to both problems mentioned above. With a little abuse of notation, we use $\Omega$ to denote either the full index set $\Omega=[d_1]\times \cdots \times [d_K]$ (for the tensor denoising) or a random subset induced from the sampling distribution $\Pi$ (for the tensor completion). Define $b_0=-\infty$, $b_L=\infty$, $f(-\infty)=0$ and $f(\infty)=1$. The log-likelihood associated with the observed entries is
\begin{align}\label{eq:objective}
\logl(\Theta, \mb)&=\sum_{\omega\in\Omega}\sum_{\ell\in[L]} \Big\{\mathds{1}_{\{y_\omega=\ell\}} \log \big[f(b_\ell-\theta_\omega)- \notag \\
&\quad \quad \quad \quad \quad \quad \quad  f(b_{\ell-1}-\theta_\omega)\big]\Big\}.
\end{align}
We propose a rank-constrained maximum likelihood estimator (a.k.a.\ M-estimator) for $\Theta$:
\begin{align}\label{eq:estimator}
\hat \Theta&=\argmax_{\Theta\in \tP}\logl(\Theta, \mb),\ \text{where}\notag \\
\tP&=\left\{\Theta\in\mathbb{R}^{d_1\times \cdots \times d_K} \colon \text{rank}(\Theta)\leq \mr,\ \mnormSize{}{\Theta}\leq \alpha\right\}.
 \end{align}
In practice, the cut-off points $\mb$ are unknown and should be jointly estimated with $\Theta$. For technical convenience, we assume in this section that the cut-off points $\mb$ are known. The adaptation of unknown $\mb$ is addressed in Section~\ref{sec:algorithm} and the Supplement.

We define a few key quantities that will be used in our theory. Let $g_\ell=f(\theta+b_\ell)-f(\theta+b_{\ell-1})$ for all $\ell\in[L]$, and
\begin{align}\label{eq:regular}
A_\alpha &= \min_{\ell\in[L], |\theta|\leq \alpha} g_\ell(\theta),\quad  U_\alpha=\max_{\ell\in[L], |\theta|\leq \alpha } {\dot{g}_\ell(\theta)\over g_\ell(\theta)},\\
 L_\alpha &= \min_{\ell\in[L], |\theta|\leq \alpha }\left[{\dot{g}_\ell^2 (\theta)\over g_\ell^2(\theta)} -{\ddot{g}_\ell(\theta)\over g_\ell(\theta)}\right],
\end{align}
where $\dot{g}(\theta)=dg(\theta)/d\theta$, and $\alpha$ is the entrywise bound of $\Theta$. In view of equation~\eqref{eq:latent}, these quantities characterize the geometry including flatness and convexity of the latent noise distribution. Under the Assumption~\ref{ass:link}, all these quantities are strictly positive and independent of tensor dimension.

\subsection{Estimation error for tensor denoising}\label{sec:denosing}
For the tensor denoising problem, we assume that the full set of tensor entries are observed. We assess the estimation accuracy using the mean squared error (MSE):
\[
\text{MSE}(\hat \Theta, \trueT)={1\over \prod_k d_k}\FnormSize{}{\Theta-\trueT}^2.
\]
The next theorem establishes the upper bound for the MSE of the proposed $\hat \Theta$ in~\eqref{eq:estimator}.

\begin{thm}[Statistical convergence] \label{thm:rate}
Consider an ordinal tensor $\tY\in[L]^{d_1\times\dots\times d_K}$ generated from model~\eqref{eq:model}, with the link function $f$ and the true coefficient tensor $\trueT\in\tP$. Define $r_{\max}=\max_k r_k$. Then, with very high probability, the estimator in~\eqref{eq:estimator} satisfies
\begin{equation}\label{eq:rate}
\text{MSE}(\hat \Theta, \trueT) \leq \min\left( 4\alpha^2,\ {c_2  U^2_\alpha r_{\max}^{K-1}  \over  L^2_\alpha } {\sum_kd_k\over  \prod_k d_k} \right),
\end{equation}
where $c_1, c_2>0$ are two constants that depend only on $K$.
\end{thm}
Theorem~\ref{thm:rate} establishes the statistical convergence for the estimator~\eqref{eq:estimator}. In fact, the proof of this theorem (see the Supplement) shows that the same statistical rate holds, not only for the global optimizer~\eqref{eq:estimator}, but also for any local optimizer $\check \Theta$ in the level set $\{\check\Theta\in\tP: \logl(\check\Theta)\geq \logl(\trueT)\}$. This suggests that the local optimality itself is not necessarily a severe concern in our context, as long as the convergent objective is large enough. In Section ~\ref{sec:algorithm}, we perform empirical studies to assess the algorithmic stability.

A similar conclusion is obtained for the prediction error, measured in Kullback-Leibler (KL) divergence, between the categorical distributions in the observation space.
\begin{cor}[Prediction error]~\label{cor:prediction}
Assume the same set-up as in Theorem~\ref{thm:rate}. Let $\mathbb{P}_{\tY}$ and $\hat{\mathbb{P}}_{\tY}$ denote the distributions generating the $L$-level ordinal tensor $\tY$, given the true parameter $\Theta$ and its estimator $\hat \Theta$, respectively. Assume $L\geq 2$. Then, with very high probability,
\begin{equation}\label{eq:KLrate}
\text{KL}(\mathbb{P}_{\tY} || \hat{\mathbb{P}}_{\tY}) \leq  {c_2 U^2_\alpha r_{\max}^{K-1}  \over L^2_\alpha } {(4L-6)\dot{f}^2(0)  \over A_\alpha} {\sum_kd_k\over  \prod_k d_k},
\end{equation}
where $c_1, c_2>0$ are the same constants as in Theorem~\ref{thm:rate}.
\end{cor}
To gain insight into these bounds, we consider a special setting with equal dimension in all modes, i.e., $d_1=\cdots=d_K=d$. In such a case, our bound \eqref{eq:rate} reduces to
\begin{equation*}\label{eq:ours}
\text{MSE}(\hat \Theta, \trueT) \asymp d^{-(K-1)}, \quad \text{as}\ d\to \infty.
\end{equation*}
Hence, our estimator achieves consistency with polynomial convergence rate. We compare the bound with existing literature. In the special case $L=2$, \citet{ghadermarzy2018learning} proposed a max-norm constrained estimator $\tilde{\Theta}$ with $\text{MSE}(\tilde{\Theta}, \trueT) \asymp  d^{-(K-1)/2}$. In contrast, our estimator converges at a rate of $d^{-(K-1)}$, which is substantially faster than theirs. This provides a positive answer to the open question posed in~\citet{ghadermarzy2018learning} whether the square root in the bound is removable. The improvement stems from utilizing the exact low-rankness of $\Theta$, whereas the surrogate rank measure employed in~\citet{ghadermarzy2018learning} is scale-sensitive.

@@

\begin{thm}[Minimax lower bound]\label{thm:minimax}
Assume the same set-up as in Theorem~\ref{thm:rate}, and $d_{\max}=\max_k d_k \geq 8$. Let $\inf_{\hat \Theta}$ denote the infimum over all estimators $\hat \Theta\in\tP$ based on the ordinal tensor observation $\tY\in[L]^{d_1\times \cdots \times d_K}$. Then, under the model~\eqref{eq:model},
\begin{align}\label{eq:lower}
\inf_{\hat \Theta }\sup_{\trueT\in\tP} \mathbb{P}\Big\{& \text{MSE}(\hat \Theta, \trueT) \\
& \geq {1\over 256} \min\left( \alpha^2, \ { Cr_{\max}d_{\max} \over \prod_k d_k} \right) \Big\} \geq {1\over 8},
\end{align}
where $C=C(\alpha, L, f,\mb)>0$ is a constant independent of tensor dimension and the rank.
\end{thm}
We see that the lower bound matches the upper bound in~\eqref{eq:rate} on the polynomial order of tensor dimension. Therefore, our estimator~\eqref{eq:estimator} is order-optimal.

@@

\vspace{-.2cm}
\section{Numerical Implementation}\label{sec:algorithm}
We describe the algorithm to seek the optimizer of~\eqref{eq:objective}. In practice, the cut-off points $\mb$ are often unknown, so we choose to maximize $\logl$ jointly over $(\Theta,\mb)\in\tP\times \tB$. The objective $\logl$ is concave in $(\Theta,\mb)$ whenever $f'$ is log-concave (see the Supplement). However, the feasible set $\tP$ is non-convex, which makes the optimization~\eqref{eq:objective} a non-convex problem. We employ the alternating optimization approach by utilizing the Tucker representation of $\Theta$. Specifically, based on~\eqref{eq:Tucker} and~\eqref{eq:objective}, the objective function consists of $K+2$ blocks of variables, one for the cut-off points $\mb$, one for the core tensor $\tC$, and $K$ for the factor matrices $\mM_k$'s. The optimization is a simple convex problem if any $K+1$ out of the $K+2$ blocks are fixed. We update one block at a time while holding others fixed, and alternate the optimization throughout the iteration. The convergence is guaranteed whenever $\logl$ is bounded from above, since the alternating procedure monotonically increases the objective. The Algorithm~\ref{alg} gives the full description.

@@

\newpage
\bibliography{tensor_wang}
\bibliographystyle{icml2020}

>>>>

\begin{icmlauthorlist}
\icmlauthor{}{}
\end{icmlauthorlist}

\icmlaffiliation{}{}
\icmlcorrespondingauthor{}{}
% You may provide any keywords that you
% find helpful for describing your paper; these are used to populate
% the "keywords" metadata in the PDF but will not be shown in the document
\icmlkeywords{Higher-order tensors, ordinal observation, tensor decomposition, tensor completion}

@@

\section{Introduction}
Multidimensional arrays, a.k.a.\ tensors, arise in a variety of applications including recommendation systems~\citep{baltrunas2011incarmusic}, social networks~\citep{nickel2011three}, genomics~\citep{hore2016tensor}, and neuroimaging~\citep{zhou2013tensor}. There is a growing need to develop general methods for analyzing these noisy, high-dimensional datasets that can handle two main problems. The problem of tensor denoising -- which aims to recover a signal tensor from its noisy entries -- has gained increased attention in theory and applications~\citep{xia2019sup,zeng2019multiway}. The second problem, tensor completion, examines the minimum number of entries needed for a consistent recovery~\citep{ghadermarzy2018learning,ghadermarzy2019near}. Low-rankness is often imposed to the signal tensor, Thereby efficiently reducing the intrinsic dimension in both problems.

@@

\vspace{-.2cm}
\section{Preliminaries}
\vspace{-.1cm}
Let $\tY\in\mathbb{R}^{d_1\times \cdots \times d_K}$ denote an order-$K$ $(d_1,\ldots,d_K)$-dimensional tensor. We use $y_\omega$ to denote the tensor entry indexed by $\omega$, where $\omega\in[d_1]\times\cdots\times[d_K]$.  The Frobenius norm of $\tY$ is defined as $\FnormSize{}{\tY}=\sum_\omega y^2_\omega$ and the infinity norm of $\tY$ is defined as $\mnormSize{}{\tY}=\max_{\omega}|y_\omega|$. We use $\tY_{(k)}$ to denote the unfolded matrix of size $d_k$-by-$\prod_{i\neq k}d_k$, obtained by reshaping the tensor along the mode-$k$. The Tucker rank of $\tY$ is defined as a length-$K$ vector $\mr = (r_1,\ldots,r_K)$, where $r_k$ is the rank of matrix $\tY_{(k)}$ for all $k \in[K]$. We say that an event $A$ occurs ``with very high probability'' if $\mathbb{P}(A)$ tends to 1 faster than any polynomial of tensor dimension $d_{\min}=\min\{d_1,\ldots,d_K\} \to\infty$.

@@

\vspace{-.2cm}
\section{Model formulation and motivation}
\vspace{-.1cm}
\subsection{Observation model}
\vspace{-.1cm}
Let $\tY$ denote an order-$K$ $(d_1,\ldots,d_K)$-dimensional data tensor. Suppose the entries of $\tY$ are ordinal-valued, and the observation space consists of $L$ ordered levels, denoted by $[L]:=\{1,\ldots,L\}$. We propose a cumulative link model for the ordinal tensor $\tY=\entry{y_\omega}\in[L]^{d_1\times \cdots\times d_K}$. Specifically, assume the entries $y_\omega$ are (conditionally) independently distributed with cumulative probabilities:
\begin{equation}\label{eq:model}
\mathbb{P}(y_\omega\leq \ell)=f(b_\ell-\theta_\omega),\ \text{for all}\ \ell\in[L-1],
\end{equation}
where $\mb=(b_1,\ldots,b_{L-1})$ is a set of unknown scalars satisfying $b_1<\cdots <b_{L-1}$, $\Theta=\entry{\theta_\omega}\in\mathbb{R}^{d_1\times \cdots \times d_K}$ is a continuous-valued parameter tensor satisfying certain low-dimensional structure (to be specified later), and $f(\cdot):\mathbb{R}\mapsto[0,1]$ is a known, strictly increasing function. We refer to $\mb$ as the cut-off points and $f$ the link function.

The formulation~\eqref{eq:model} imposes an additive model to the transformed probability of cumulative categories. This modeling choice is to respect the ordering structure among the categories. For example, if we choose the inverse link $f^{-1}(x)=\log {x\over 1-x}$ to be the log odds, then the model~\eqref{eq:model} implies linear spacing between the proportional odds:
\begin{equation}\label{eq:logodd}
\log {\mathbb{P}(y_{\omega}\leq \ell) \over \mathbb{P}(y_\omega >  \ell) } - \log {\mathbb{P}(y_{\omega}\leq {\ell-1}) \over \mathbb{P}(y_\omega >  {\ell-1}) } = b_\ell-b_{\ell-1},
\end{equation}
for all tensor entries $y_\omega$. When there are only two categories in the observation space (e.g. binary tensors), the cumulative model~\eqref{eq:model} is equivalent to the usual multinomial link model. In general, however, when the number of categories $L\geq 3$, the proportional odds assumption~\eqref{eq:logodd} is more parsimonious, in that, the ordered categories can be envisaged as contiguous intervals on the continuous scale, where the points of division are exactly $b_1<\cdots <b_{L-1}$. This interpretation will be made more explicit in the next section.
\vspace{-.1cm}
\subsection{Latent-variable interpretation}\label{sec:latent}
\vspace{-.1cm}

The ordinal tensor model~\eqref{eq:model} with certain types of link $f$ has the equivalent representation as an $L$-level quantization model on $\tY=\entry{y_\omega}$:
\begin{align}\label{eq:quantization}
y_\omega&=
\begin{cases}
1,& \text{if $y^*_\omega\in(-\infty, b_1]$},\\
2,& \text{if $y^*_\omega\in(b_1, b_2]$},\\
\vdots  &\vdots\\
L,& \text{if $y^*_\omega\in(b_{L-1}, \infty)$},\\
\end{cases}
\end{align}
for all $\omega\in[d_1]\times \cdots \times [d_k]$. Here, $\tY^*=\entry{y^*_\omega}$ is a latent continuous-valued tensor following an additive noise model:
\begin{align}\label{eq:latent}
\KeepStyleUnderBrace{\tY^*}_{\text{latent continuous-valued tensor}}=\KeepStyleUnderBrace{\Theta}_{\text{signal tensor}}+\KeepStyleUnderBrace{\tE}_{\text{i.i.d.\ noise}},
\end{align}
where $\tE=\entry{\varepsilon_\omega}\in\mathbb{R}^{d_1\times \cdots \times d_K}$ is a noise tensor with i.i.d.\ entries according to distribution $\mathbb{P}(\varepsilon)$. From the viewpoint of~\eqref{eq:latent}, the parameter tensor $\Theta$ can be interpreted as the latent signal tensor prior to contamination and quantization.

@@

\begin{assumption}\label{ass:link}
The link function is assumed to satisfy:
\vspace{-.4cm}
\begin{enumerate}
\item $f(\theta)$ is strictly increasing and twice-differentiable in $\theta\in\mathbb{R}/\{0\}$.
\vspace{-.2cm}
\item $f'(\theta)$ is strictly log-concave and symmetric with respect to $\theta=0$.
\end{enumerate}
\end{assumption}
\vspace{-.1cm}
\subsection{Problem 1: Tensor denoising}~\label{sec:denoising}
\vspace{-.1cm}
The first question we aim to address is tensor denoising:

@@

The second constraint is that the entries of $\Theta$ are uniformly bounded in magnitude by a constant $\alpha \in \mathbb{R}_{+}$. In view of~\eqref{eq:latent}, we refer to $\alpha$ as the signal level. The entry-wise bound assumption is a technical condition that avoids the degeneracy in probability estimation with ordinal observations.
\vspace{-.1cm}
\subsection{Problem 2: Tensor completion}
\vspace{-.1cm}
Motivated by applications in collaborative filtering, we also consider a more general setup when only a subset of tensor entries $y_\omega$ are observed. Let $\Omega\subset[d_1]\times \cdots\times[d_K]$ denote the set of observed indices. The second question we aim to address is stated as follows:

@@

We consider the same parameter space~\eqref{eq:space} for the completion problem. In addition to the reasons mentioned in Section~\ref{sec:denoising}, the entrywise bound assumption also serves as the incoherence requirement for completion. In classical matrix completion, the incoherence is often imposed on the singular vectors. This assumption is recently relaxed for ``flat'' matrices with bounded magnitude~\cite{negahban2011estimation,cai2013max,bhaskar20151}. We adopt the same assumption for higher-order tensors.
\vspace{-.2cm}
\section{Rank-constrained M-estimator}\label{sec:theory}
\vspace{-.1cm}
We present a general treatment to both problems mentioned above. With a little abuse of notation, we use $\Omega$ to denote either the full index set $\Omega=[d_1]\times \cdots \times [d_K]$ (for the tensor denoising) or a random subset induced from the sampling distribution $\Pi$ (for the tensor completion). Define $b_0=-\infty$, $b_L=\infty$, $f(-\infty)=0$ and $f(\infty)=1$. The log-likelihood associated with the observed entries is
\begin{align}\label{eq:objective}
\logl(\Theta, \mb)&=\sum_{\omega\in\Omega}\sum_{\ell\in[L]} \Big\{\mathds{1}_{\{y_\omega=\ell\}} \log \big[f(b_\ell-\theta_\omega)- \notag \\
&\quad \quad \quad \quad \quad \quad \quad  f(b_{\ell-1}-\theta_\omega)\big]\Big\}.
\end{align}
We propose a rank-constrained maximum likelihood estimator (a.k.a.\ M-estimator) for $\Theta$:
\begin{align}\label{eq:estimator}
\hat \Theta&=\argmax_{\Theta\in \tP}\logl(\Theta, \mb),\ \text{where}\notag \\
\tP&=\left\{\Theta\in\mathbb{R}^{d_1\times \cdots \times d_K} \colon \text{rank}(\Theta)\leq \mr,\ \mnormSize{}{\Theta}\leq \alpha\right\}.
 \end{align}
In practice, the cut-off points $\mb$ are unknown and should be jointly estimated with $\Theta$. For technical convenience, we assume in this section that the cut-off points $\mb$ are known. The adaptation of unknown $\mb$ is addressed in Section~\ref{sec:algorithm} and the Supplement.

We define a few key quantities that will be used in our theory. Let $g_\ell=f(\theta+b_\ell)-f(\theta+b_{\ell-1})$ for all $\ell\in[L]$, and
\begin{align}\label{eq:regular}
A_\alpha &= \min_{\ell\in[L], |\theta|\leq \alpha} g_\ell(\theta),\quad  U_\alpha=\max_{\ell\in[L], |\theta|\leq \alpha } {|\dot{g}_\ell(\theta)|\over g_\ell(\theta)},\\
 L_\alpha &= \min_{\ell\in[L], |\theta|\leq \alpha }\left[{\dot{g}_\ell^2 (\theta)\over g_\ell^2(\theta)} -{\ddot{g}_\ell(\theta)\over g_\ell(\theta)}\right],
\end{align}
where $\dot{g}(\theta)=dg(\theta)/d\theta$, and $\alpha$ is the entrywise bound of $\Theta$. In view of equation~\eqref{eq:latent}, these quantities characterize the geometry including flatness and convexity of the latent noise distribution. Under the Assumption~\ref{ass:link}, all these quantities are strictly positive and independent of tensor dimension.
\vspace{-.1cm}
\subsection{Estimation error for tensor denoising}\label{sec:denosing}
\vspace{-.1cm}
For the tensor denoising problem, we assume that the full set of tensor entries are observed. We assess the estimation accuracy using the mean squared error (MSE):
\[
\text{MSE}(\hat \Theta, \trueT)={1\over \prod_k d_k}\FnormSize{}{\Theta-\trueT}^2.
\]
The next theorem establishes the upper bound for the MSE of the proposed $\hat \Theta$ in~\eqref{eq:estimator}.

\begin{thm}[Statistical convergence] \label{thm:rate}
Consider an ordinal tensor $\tY\in[L]^{d_1\times\dots\times d_K}$ generated from model~\eqref{eq:model}, with the link function $f$ and the true coefficient tensor $\trueT\in\tP$. Define $r_{\max}=\max_k r_k$. Then, with very high probability, the estimator in~\eqref{eq:estimator} satisfies
\begin{equation}\label{eq:rate}
\mathrm{MSE}(\hat \Theta, \trueT) \leq \min\left( 4\alpha^2,\ {c_1  U^2_\alpha r_{\max}^{K-1}  \over  L^2_\alpha } {\sum_kd_k\over  \prod_k d_k} \right),
\end{equation}
where $c_1 >0$ is a constant that depends only on $K$.
\end{thm}
Theorem~\ref{thm:rate} establishes the statistical convergence for the estimator~\eqref{eq:estimator}. In fact, the proof of this theorem (see the Supplement) shows that the same statistical rate holds, not only for the global optimizer~\eqref{eq:estimator}, but also for any local optimizer $\check \Theta$ in the level set $\{\check\Theta\in\tP: \logl(\check\Theta)\geq \logl(\trueT)\}$. This suggests that the local optimality itself is not necessarily a severe concern in our context, as long as the convergent objective is large enough. In Section ~\ref{sec:algorithm}, we perform empirical studies to assess the algorithmic stability.

A similar conclusion is obtained for the prediction error, measured in Kullback-Leibler (KL) divergence, between the categorical distributions in the observation space.
\begin{cor}[Prediction error]~\label{cor:prediction}
Assume the same set-up as in Theorem~\ref{thm:rate}. Let $\mathbb{P}_{\tY}$ and $\hat{\mathbb{P}}_{\tY}$ denote the distributions generating the $L$-level ordinal tensor $\tY$, given the true parameter $\Theta$ and its estimator $\hat \Theta$, respectively. Assume $L\geq 2$. Then, with very high probability,
\begin{equation}\label{eq:KLrate}
\text{KL}(\mathbb{P}_{\tY} || \hat{\mathbb{P}}_{\tY}) \leq  {c_1 U^2_\alpha r_{\max}^{K-1}  \over L^2_\alpha } {(4L-6)\dot{f}^2(0)  \over A_\alpha} {\sum_kd_k\over  \prod_k d_k},
\end{equation}
where $c_1 >0$ is the same constants as in Theorem~\ref{thm:rate}.
\end{cor}
To gain insight into these bounds, we consider a special setting with equal dimension in all modes, i.e., $d_1=\cdots=d_K=d$. In such a case, our bound \eqref{eq:rate} reduces to
\begin{equation}\label{eq:ours}
\text{MSE}(\hat \Theta, \trueT) \asymp d^{-(K-1)}, \quad \text{as}\ d\to \infty.
\end{equation}
Hence, our estimator achieves consistency with polynomial convergence rate. We compare the bound with existing literature. In the special case $L=2$, \citet{ghadermarzy2018learning} proposed a max-norm constrained estimator $\tilde{\Theta}$ with $\text{MSE}(\tilde{\Theta}, \trueT) \asymp  d^{-(K-1)/2}$. In contrast, our estimator converges at a rate of $d^{-(K-1)}$, which is substantially faster than theirs. This provides a positive answer to the open question posed in~\citet{ghadermarzy2018learning} whether the square root in the bound is removable. The improvement stems from utilizing the exact low-rankness of $\Theta$, whereas the surrogate rank measure employed in~\citet{ghadermarzy2018learning} is scale-sensitive.

@@

\begin{thm}[Minimax lower bound]\label{thm:minimax}
Assume the same set-up as in Theorem~\ref{thm:rate}, and $d_{\max}=\max_k d_k \geq 8$. Let $\inf_{\hat \Theta}$ denote the infimum over all estimators $\hat \Theta\in\tP$ based on the ordinal tensor observation $\tY\in[L]^{d_1\times \cdots \times d_K}$. Then, under the model~\eqref{eq:model},
\begin{align}\label{eq:lower}
\inf_{\hat \Theta }\sup_{\trueT\in\tP} \mathbb{P}\Big\{& \text{MSE}(\hat \Theta, \trueT) \\
& \geq c\min\left( \alpha^2, \ { Cr_{\max}d_{\max} \over \prod_k d_k} \right) \Big\} \geq {1\over 8},
\end{align}
where $C=C(\alpha, L, f,\mb)>0$ and $c>0$ are constants independent of tensor dimension and the rank.
\end{thm}
We see that the lower bound matches the upper bound in~\eqref{eq:rate} on the polynomial order of tensor dimension. Therefore, our estimator~\eqref{eq:estimator} is order-optimal.

@@

\vspace{-.2cm}
\section{Numerical Implementation}\label{sec:algorithm}
\vspace{-.1cm}
We describe the algorithm to seek the optimizer of~\eqref{eq:objective}. In practice, the cut-off points $\mb$ are often unknown, so we choose to maximize $\logl$ jointly over $(\Theta,\mb)\in\tP\times \tB$. The objective $\logl$ is concave in $(\Theta,\mb)$ whenever $f'$ is log-concave (see the Supplement). However, the feasible set $\tP$ is non-convex, which makes the optimization~\eqref{eq:objective} a non-convex problem. We employ the alternating optimization approach by utilizing the Tucker representation of $\Theta$. Specifically, based on~\eqref{eq:Tucker} and~\eqref{eq:objective}, the objective function consists of $K+2$ blocks of variables, one for the cut-off points $\mb$, one for the core tensor $\tC$, and $K$ for the factor matrices $\mM_k$'s. The optimization is a simple convex problem if any $K+1$ out of the $K+2$ blocks are fixed. We update one block at a time while holding others fixed, and alternate the optimization throughout the iteration. The convergence is guaranteed whenever $\logl$ is bounded from above, since the alternating procedure monotonically increases the objective. The Algorithm~\ref{alg} gives the full description.

@@

\newpage
\bibliography{ordinalTv3.bbl}
\bibliographystyle{icml2020}
